Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-11 write-ups/PLDI/paper.tex

\paragraph{A degree of fairness: occupancy-bound execution}
GPU schedulers are \emph{not} fair. the current GPU
programming models---OpenCL~\cite{opencl2Spec}, CUDA~\cite{...}  and
HSA~\cite{HSAprogramming11}---specify almost no guarantees regarding scheduling of
workgroups, and implementations of these programming models do not
provide fair scheduling in practice.  Roughly speaking, each workgroup
executing a GPU kernel is mapped to a hardware \emph{compute
  unit}.\footnote{In practice the situation can be more complex: if a
  kernel is sufficiently simple and if workgroups are small enough,
  multiple workgroups may be mapped to a single compute unit.  We assume, for the purpose of the current discussion, that the kernel is sufficiently complex that this does not happen.}
%
If a kernel is executed with more workgroups than there are compute
units then at least two workgroups must share a single compute unit.
The simplest way for a GPU driver to handle this is via an \emph{occupancy-bound}
execution model~\cite{...}.  With this model, once a workgroup has commenced execution on a
compute unit (it has become \emph{occupant}), the workgroup has
exclusive access to the compute unit until it has finished execution,
after which a subsequent workgroup can be bound to the compute unit.
Experiments suggest that the occupancy-bound execution model is widely
implemented by today's GPU devices and drivers~\cite{...}.  It is
evidently a simple model for device vendors to implement efficiently.

The occupancy-bound execution model does not guarantee fair scheduling
between workgroups: if all compute units are occupied then a
not-yet-occupant workgroup will not be scheduled until some occupant
workgroup completes execution.  Yet the execution model \emph{does}
provide fair scheduling between \emph{occupant} workgroups, because
they are bound to separate compute units that operate in parallel.
Current implementations of irregular data parallel algorithms on GPUs
assume that the target GPU device provides the occupancy-bound
execution model.  They exploit the model by scheduling no more workgroups
than there are available compute units~\cite{...}.  This works
because \emph{today}'s GPUs seem to provide the occupancy-bound
execution model.

@@

Despite its practical prevalence, none of the current GPU programming
models actually mandate occupancy-bound execution.  Further, there are
clear reasons why this model is undesirable from the point-of-view of
GPU vendors and end users.  First, the execution model does not enable
multi-tasking, because a workgroup effectively \emph{owns} a compute
unit until the workgroup has completed execution.  If a long-running
kernel occupies all compute units, the GPU cannot be used for other
tasks, such as graphics rendering, in the meantime.  Precisely because
of this problem, operating systems employ a GPU \emph{watchdog} to
heuristically detect and kill long-running computations on the GPU
being used for display rendering~\cite{...}.  Second, \emph{energy throttling} is
an important concern for battery-powered devices, such as smartphones
and tablets~\cite{...}.  In the future, it will be desirable for a mobile GPU driver to be able to
power down one or more compute units if the battery level is low or if
the user puts the device into a low power state.  Implementing energy
throttling would require being able to suspend the execution of a
workgroup, power down the compute unit to which the workgroup was
bound, and schedule the workgroup for later execution on a different
compute unit.

Our assessment, informed by discussions with a number of industrial
practitioners who have been involved in the OpenCL and/or HSA
standardisation efforts~\cite{CitePersonalCommunication}, is that (1)
GPU vendors will not commit to the occupancy-bound execution model
that they currently implement, due to the need for multi-tasking and
energy throttling, yet (2) GPU vendors will not guarantee fair
scheduling, due to the runtime overhead associated with supporting
full preemption of workgroups~\cite{ISCAPAPERSreeMentioned}.  Vendors
instead wish to retain the essence of the occupancy-bound execution
model, which is simple to implement efficiently, and provide
support for preemption only in key special cases.

@@

Fine-grained communication between threads, and also with the host in
the case of SVM, is enabled via a set of atomic data types and
operations, based on the C11 memory model~\cite{...}.  \ADComment{Could cut:} In particular,
atomic SVM operations enable fine-grained communication between host
and device.  Our implementation of cooperative kernels
(\mysec\ref{sec:implementation}) depends on this recent OpenCL
feature.

@@

\paragraph{Work stealing example}
%
The work stealing paradigm is useful when a computation can be split
into multiple tasks that can be processed independently, but where the
number of tasks is not known ahead of time because processing a given
task may create an arbitrary number of new tasks. Work stealing enables
dynamic balancing of tasks across several processing units, OpenCL
workgroups in our case. Each workgroup has an associated task pool from
which it obtains tasks to process, and to which it stores newly created
tasks. When a workgroup's pool becomes empty, a workgroup---rather than
terminating---tries to \emph{steal} a task from an other
pool. Workgroups keep on processing tasks until all the work is done.

@@

\paragraph{Graph traversal example} \myfiglong\ref{fig:graphsearch} shows the
skeleton of a frontier-based graph traversal algorithm; such algorithms have
been shown to be execute efficiently on GPUs~\cite{...}.
The kernel is
given three arguments in global memory: a graph structure, and two
arrays that can contain graph nodes. Initially, $\keyword{n0}$ contains the
starting nodes to process. Threads keep three local variables: $\keyword{level}$, which indicates the frontier level currently being
processed, and $\keyword{in\_nodes}$ and $\keyword{out\_nodes}$, which point to
distinct node arrays. The nodes to be
processed for the current frontier are contained in $\keyword{in\_nodes}$, while $\keyword{out\_nodes}$ contains
the nodes to be processed during the next frontier.

@@

The global barrier used in this application is not provided as a GPU
primitive, though many previous works have shown that such a global
barrier can be implemented~\cite{...}. These implementations use well
defined atomic memory operations to ensure memory consistency, and
employ spinning to ensure threads wait at the barrier until all
threads have arrived.  This spinning behaviour requires fair
scheduling between workgroups for the barrier to operate correctly,
and as a result the barrier is \emph{not} guaranteed to work correctly
under current GPU programming models.  Such barriers appear to run
reliably on current GPUs for kernels that are executed with no more
workgroups than there are compute units; this is due to the fairness
of the occupancy-bound execution model that current GPUs appear to
provide.  But, as discussed in \mysec\ref{sec:intro}, this
occupancy-bound model is not mandated by language standards or vendor
implementations, and is unlikely to be guaranteed in the future.

@@

Our $\globalbarrier$ primitive is a kernel-level function: if it
appears in conditional code then it must be reached by \emph{all}
threads executing the cooperative kernel.  On reaching a
$\globalbarrier()$, a thread waits until all threads have arrived at
the barrier.  Once all threads have arrived, the threads may proceed
past the barrier with the guarantee that all global memory accesses
issued before the barrier have completed.  The $\globalbarrier$
primitive can be implemented following any inter-workgroup barrier
design, \TSComment{This isn't exactly true. For cooperative kernels,
  the barriers must be able to handle a growing or shrinking number of
  workgroups.} e.g.~\cite{...}, and the atomic operations provided by
the OpenCL 2.0 memory model enable a memory-safe
implementation~\cite{...}.  However, implementing such a barrier is
involved, hence why we include this function as a primitive.

@@

\section{Notes}

\ADComment{A natural exit does not decrement $M$.  We haven't made a point of this, and I don't think we should right now, but we should remember to think this through carefully.}

  \ADComment{Perhaps
  make the point that inter-workgroup synchronization is possible in
  very rare cases without blocking, e.g. threadfence reduction.}

\ADComment{If this watchdog is disabled,
execution of a long-running kernel causes the display to freeze for
the duration of kernel execution.}

\section{Remainder of old intro}

In an effort to achieve high performance for IP applications, new GPU
programming paradigms have been developed that include custom GPU
synchronisation constructs. Here we focus on two \ADComment{The way this is worded it is unclear whether you are describing two programming paradigms, or two synchronisation constructs}:

\begin{enumerate}
\item {\bf Inter-workgroup barrier}: Workgroups are able to
  efficiently synchronise frequently using a custom inter-workgroup
  barrier that is not provided as GPU primitive.

\item {\bf Work stealing}: Computation is divided into tasks that are
  stored in workgroup-local queues. If a workgroup runs out of local
  tasks to compute, it is able to steal a task from a different
  workgroup.
\end{enumerate}

Both paradigms fit into what are called \emph{persistent kernels}, the
defining characteristic being that persistent kernels use a relatively
small number of long-running threads to perform computation. In
contrast, traditional GPU programs employ many short running threads
to perform computation. In this work, the persistent kernels we consider
use either work stealing or inter-workgroup barrier paradigms.

Persistent kernels rely on strong forward progress properties (FPPs)
between workgroups; that is, workgroups must make forward progress with respect to
the other workgroups that are executing. In the case of the inter-workgroup barrier, all
workgroups spin at a barrier instruction waiting for all workgroups to
arrive. If one (or more) workgroups stop making forward progress, the
application will deadlock as the live workgroups will wait
indefinitely for the halted workgroups \ADComment{Can we re-word the preceding to make it clearer that the workgroups have been stalled by a scheduler: they have not actually terminated their execution by reaching the end of the kernel}. Less obviously, in the work
stealing paradigm, the workgroups terminate only when \emph{all} tasks
have been completed. If a workgroup takes a task, but stops making
forward progress before the task is completed, the application will
deadlock as the live workgroups will spin waiting for the halted
workgroup to finish its task.

The current GPU programming models generally do not provide strong
enough inter-workgroup FPP to allow persistent kernels. \ADComment{This might be a good place to cite the specs.}  This is
because GPUs do not provide workgroup preemption. That is, GPU
programming models allow GPU programs to be launched with many more
workgroups than the available hardware resources are able run
concurrently. The GPU runtime will schedule as many workgroups as the
available resources can provide, and the remaining workgroups will
wait until hardware resources become available, e.g.  when one of the
scheduled workgroups finishes execution. There is no mechanism to swap
workgroups from 'executing' to 'waiting' and vice versa. Workgroups
that have been scheduled and have not finished execution are called
\emph{occupant}.  \ADComment{Note: it's ironic that lack of preemption is what leads to the problem of workgroups being scheduled in waves, and yet also what means we might not be able to guarantee the properties of the occupancy-bound execution model in the future.}

Previous work experimenting with persistent kernels has required
\emph{a priori} knowledge of the number of workgroups that the GPU
hardware is able to execute the program with concurrently. This
number, called the \emph{maximal launch} or \emph{occupancy bound}, is
a function of both a GPU chip and a compiled program. Because occupant
workgroups are never preempted, if the GPU program is launched with a
number of workgroups which doesn't exceed the occupancy bound, then
all workgroups will be occupant and the strong inter-workgroup FPP
required by persistent kernels are provided. Using the occupancy
bound, previous work has successfully implemented persistent kernels
and shown them to be performant on current GPUs. Recent work has even
shown that the occupancy bound can be accurately discovered
programitically, removing the need for any \emph{a priori} knowledge
about the GPU.

Despite the promising performance and ease of programming persistent
kernels with occupant workgroups, GPU language specification
committees are reluctant to commit to the inter-workgroup FPP
currently provided by occupant workgroups. That is, even if current
GPUs provide strong FPP between workgroups, future GPUs may not. Thus
the performance gains enabled by persistent kernels may not be
safely achievable for future GPUs.

We believe that two fundamental issues stopping committees from
committing to strong inter-workgroup FPP are:

\begin{itemize}
\item {\bf Responsive GPU multi-tasking} A high priority GPU task
  (e.g. graphics) may need to quickly gain access to the GPU. To do
  this, some of the resources will have to be taken from the
  persistent kernel in order to run the high priority task. This will
  reduce the number of occupant workgroups executing the persistent
  kernel.

\item {\bf Energy throttling} Some GPUs, especially in mobile devices,
  may wish to monitor the energy consumed (i.e. heat output) of the
  GPU chip.  A GPU that is consuming too much energy may be throttled
  by limiting the amount of resources (and thus the number of occupant
  workgroups) to conserve battery life.
\end{itemize}

Both issues are a case of the GPU runtime needing to adjust the
resources used by the persistent kernel during execution. This causes
occupant workgroups to potentially be preempted, which may lead to
deadlock in persistent kernels. \ADComment{It might be good to illustrate the previous points with an example; e.g.\ we could envisage a kernel that is being executed concurrently by 16 workgroups.  Some (4 say) have reached an inter workgroup barrier and are awaiting the others.  But then a graphics tasks causes the 12 workgroups that have not yet reached the barrier to be preempted, and in anticipation of further graphics processing requirements the scheduler decides not to reduce the workgroup allocation for the kernel to 4 indefinitely.  The 4 workgroups waiting at the barrier cannot now make progress.  Energy throttling example is analogous, except that the count is reduced from 16 to 12 due to low battery.}
A less severe, but still undesirable
issue is if the runtime determines more GPU resources are available
during a persistent kernel execution. In this case, the persistent
kernel will be under-utilising GPU resources. \ADComment{I'm not sure I get this one.  Do you mean here that at the point of kernel launch there may be e.g. only 4 workgroup slots available, say due to low battery, but then later on (e.g., due to a phone being charged) there may be scope for more workgroups to run in parallel?}

\ADComment{We should mention that GPU watchdogs are the rather drastic solution to this problem, to keep the overall system responsive, but that they do not help from the point of view of IP applications themselves.}

In this work, we propose a solution that unites persistent kernels
with the need for GPU runtimes to dynamically (and quickly) adjust the
resources available to the persistent kernel. We propose a new GPU
programming paradigm called \emph{dynamic persistent kernels}, or
DPK. DPKs provide strong inter-workgroup FPP while still allowing the
GPU runtime to adjust the number of workgroups executing the DPK in a
timely manner, enabling responsive GPU multi-tasking or energy
throttling.

The key insight behind DPK, is that many persistent kernels contain
points in their execution where workgroups can easily be either
terminated or started. We propose several new instructions to mark
these entry and exit points. These instructions serve as hooks for the
GPU runtime to adjust the amount of resources available to the DPK,
and thus the number of occupant workgroups the DPK can rely on to have
strong FPPs. \ADComment{We should make it clear early on that some care on the part of the programmer is required to make this work, and that if the programmer does not insert sufficient yield points, the GPU watchdog may still have to kill a kernel.}

For example, many applications using the inter-workgroup barrier
paradigm do not carry workgroup-local information across calls to the
inter-workgroup barrier. Thus the number of workgroups can be adjusted
at every call to the inter-workgroup barrier. Many applications using
the work stealing paradigm do not carry state between task
computation; thus workgroups can be created or terminated immediately
after ever task computation.

While we envision support for DPKs to ultimately be implemented in
proprietary GPU drivers, we are able to provide a prototype
implementation in software using the strong inter-workgroup FPP
available on current GPUs. This allows us to test the feasibility of
our approach and as well get upper bounds for the performance cost and
the responsiveness of multi-tasking.

>>>>

\paragraph{A degree of fairness: occupancy-bound execution}
GPU schedulers are \emph{not} fair. the current GPU
programming models---OpenCL~\cite{opencl2Spec}, \ADComment{Tyler, please add best CUDA since I've asked you to add it below in any case} CUDA~\cite{...}  and
HSA~\cite{HSAprogramming11}---specify almost no guarantees regarding scheduling of
workgroups, and implementations of these programming models do not
provide fair scheduling in practice.  Roughly speaking, each workgroup
executing a GPU kernel is mapped to a hardware \emph{compute
  unit}.\footnote{In practice the situation can be more complex: if a
  kernel is sufficiently simple and if workgroups are small enough,
  multiple workgroups may be mapped to a single compute unit.  We assume, for the purpose of the current discussion, that the kernel is sufficiently complex that this does not happen.}
%
If a kernel is executed with more workgroups than there are compute
units then at least two workgroups must share a single compute unit.
The simplest way for a GPU driver to handle this is via an \emph{occupancy-bound}
execution model~\cite{DBLP:conf/oopsla/SorensenDBGR16}. \ADComment{Tyler, please add best cite for OB model; I guess our OOPSLA paper introduced the term, but we should cite that + Owens?}  With this model, once a workgroup has commenced execution on a
compute unit (it has become \emph{occupant}), the workgroup has
exclusive access to the compute unit until it has finished execution,
after which a subsequent workgroup can be bound to the compute unit.
Experiments suggest that the occupancy-bound execution model is widely
implemented by today's GPU devices and drivers~\cite{DBLP:conf/oopsla/SorensenDBGR16}. \ADComment{For the previous cite, should we cite other papers as well as ours?}  It is
evidently a simple model for device vendors to implement efficiently.

The occupancy-bound execution model does not guarantee fair scheduling
between workgroups: if all compute units are occupied then a
not-yet-occupant workgroup will not be scheduled until some occupant
workgroup completes execution.  Yet the execution model \emph{does}
provide fair scheduling between \emph{occupant} workgroups, because
they are bound to separate compute units that operate in parallel.
Current implementations of irregular data parallel algorithms on GPUs
assume that the target GPU device provides the occupancy-bound
execution model.  They exploit the model by scheduling no more workgroups
than there are available compute units~\cite{...}. \ADComment{Tyler, please pick best cite for previous claim.} This works
because \emph{today}'s GPUs seem to provide the occupancy-bound
execution model.

@@

Despite its practical prevalence, none of the current GPU programming
models actually mandate occupancy-bound execution.  Further, there are
clear reasons why this model is undesirable from the point-of-view of
GPU vendors and end users.  First, the execution model does not enable
multi-tasking, because a workgroup effectively \emph{owns} a compute
unit until the workgroup has completed execution.  If a long-running
kernel occupies all compute units, the GPU cannot be used for other
tasks, such as graphics rendering, in the meantime.  Precisely because
of this problem, operating systems employ a GPU \emph{watchdog} to
heuristically detect and kill long-running computations on the GPU
being used for display rendering~\cite{DBLP:conf/iwocl/SorensenD16}.  Second, \emph{energy throttling} is
an important concern for battery-powered devices, such as smartphones
and tablets~\cite{DBLP:journals/comsur/Vallina-RodriguezC13}.  In the future, it will be desirable for a mobile GPU driver to be able to
power down one or more compute units if the battery level is low or if
the user puts the device into a low power state.  Implementing energy
throttling would require being able to suspend the execution of a
workgroup, power down the compute unit to which the workgroup was
bound, and schedule the workgroup for later execution on a different
compute unit.

Our assessment, informed by discussions with a number of industrial
practitioners who have been involved in the OpenCL and/or HSA
standardisation efforts
(including~\cite{PersonalCommunicationRichards,PersonalCommunicationHowes}
and various engineers from GPU vendors), is that (1) GPU vendors will
not commit to the occupancy-bound execution model that they currently
implement, due to the need for multi-tasking and energy throttling,
yet (2) GPU vendors will not guarantee fair scheduling, due to the
runtime overhead associated with supporting full preemption of
workgroups~\cite{ISCAPAPERSreeMentioned}.  Vendors instead wish to
retain the essence of the occupancy-bound execution model, which is
simple to implement efficiently, and provide support for preemption
only in key special cases.

@@

Fine-grained communication between threads, and also with the host in
the case of SVM, is enabled via a set of atomic data types and
operations, based on the C11 memory model~\cite{C11}.

@@

\paragraph{Work stealing example}  \ADComment{Tyler/Hugues: can you please say in the first paragraph that this has been implemented on GPUs and cite the right papers?}
%
The work stealing paradigm is useful when a computation can be split
into multiple tasks that can be processed independently, but where the
number of tasks is not known ahead of time because processing a given
task may create an arbitrary number of new tasks. Work stealing enables
dynamic balancing of tasks across several processing units, OpenCL
workgroups in our case. Each workgroup has an associated task pool from
which it obtains tasks to process, and to which it stores newly created
tasks. When a workgroup's pool becomes empty, a workgroup---rather than
terminating---tries to \emph{steal} a task from an other
pool. Workgroups keep on processing tasks until all the work is done.

@@

\paragraph{Graph traversal example} \myfiglong\ref{fig:graphsearch} shows the
skeleton of a frontier-based graph traversal algorithm; such algorithms have
been shown to be execute efficiently on GPUs~\cite{...}. \ADComment{Tyler, can you please work out which of the graph algorithm papers is the right one to cite here?}
The kernel is
given three arguments in global memory: a graph structure, and two
arrays that can contain graph nodes. Initially, $\keyword{n0}$ contains the
starting nodes to process. Threads keep three local variables: $\keyword{level}$, which indicates the frontier level currently being
processed, and $\keyword{in\_nodes}$ and $\keyword{out\_nodes}$, which point to
distinct node arrays. The nodes to be
processed for the current frontier are contained in $\keyword{in\_nodes}$, while $\keyword{out\_nodes}$ contains
the nodes to be processed during the next frontier.

@@

The global barrier used in this application is not provided as a GPU
primitive, though previous works have shown that such a global
barrier can be implemented~\cite{XF10,DBLP:conf/oopsla/SorensenDBGR16}. \ADComment{Are there additional papers we should be citing on GPU barriers?  E.g., is the sense reversal idea part of XF or is that from another paper?}  These implementations use well
defined atomic memory operations to ensure memory consistency, and
employ spinning to ensure threads wait at the barrier until all
threads have arrived.  This spinning behaviour requires fair
scheduling between workgroups for the barrier to operate correctly,
and as a result the barrier is \emph{not} guaranteed to work correctly
under current GPU programming models.  Such barriers appear to run
reliably on current GPUs for kernels that are executed with no more
workgroups than there are compute units; this is due to the fairness
of the occupancy-bound execution model that current GPUs appear to
provide.  But, as discussed in \mysec\ref{sec:intro}, this
occupancy-bound model is not mandated by language standards or vendor
implementations, and is unlikely to be guaranteed in the future.

@@

Our $\globalbarrier$ primitive is a kernel-level function: if it
appears in conditional code then it must be reached by \emph{all}
threads executing the cooperative kernel.  On reaching a
$\globalbarrier()$, a thread waits until all threads have arrived at
the barrier.  Once all threads have arrived, the threads may proceed
past the barrier with the guarantee that all global memory accesses
issued before the barrier have completed.  The $\globalbarrier$
primitive can be implemented by adapting an inter-workgroup barrier
design, e.g.~\cite{XF}, to take account of a growing and shrinking number of workgroups, and the atomic operations provided by
the OpenCL 2.0 memory model enable a memory-safe
implementation~\cite{DBLP:conf/oopsla/SorensenDBGR16}.  However, implementing such a barrier is
involved, hence why we include this function as a primitive.

@@
